Communication-efficient Learning from Multiple Experts

### 2017-11-01 draft/draft.tex
draft update

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure} 
\usepackage{color}
%\usepackage{subfig}%[caption=false,font=footnotesize]

@@

\section{Introduction}
Supervised machine learning depends on our ability to gather a large set of labeled data. In a large number of scenarios, we can benefit from labeled data gathered by querying multiple separate sources with varying reliability and join this data into one machine learning model. However, many times we are limited in the number of data sources that we are allowed to use, because of budget constraints, limited processing power or insufficient communication bandwidth. This causes the need for a framework that would enable resource-efficient combination of labeled data from multiple sources.

First important fitting scenario to this framework is crowdsourcing. When using crowdsourcing systems such as Amazon Mechanical Turk, we can gather labelled data from multiple annotators. After gathering this training data, we need to join the labelling decisions and train one machine learning system while leveraging all of the annotations. Relying on a concept of \textit{wisdom of crowds}, we can assume that if we gather data from a sufficient number of people, on average most of them will give correct labelling decision. However, gathering data from a large amount of annotators can be expensive, as workers on Amazon Mechanical Turk require payment for their work. Therefore we would benefit from a procedure of prior determination of the annotators' expertise. In this case, the joining process can be optimized by estimating the expertise of the annotators, selecting the optimal subset and determining weights of the annotators' labels.

Moreover, in a rising area of \textit{Internet of Things} (IoT) we can have similar problems. In IoT systems, a large number of sensors provide information about the environment. Communication among multiple sensors can be costly in terms of processing power. Furthermore, this communication can be done among sensors that are limited with power consumption constraints.

There are multiple papers that deal with estimating the expertise of labelers when using \textit{wisdom of crowds}. For example, Welinder et al. \cite{welinder2010multidimensional} determine groups of annotators with similar expertise, and find particular phenomena such as "schools of thought" and specialist experts in certain types of image labeling tasks. Tian and Zhu~\cite{tian2012learning} investigate the annotator behavior in executing more complex labeling tasks, such as answering if an image is beautiful or if it contains a car. Zhang et al.~\cite{zhang2013learning} propose a method to filter out novice labelers. However, their method is specific to Gaussian Mixture Model. Rodrigues et al.~\cite{rodrigues2014gaussian} determine a similar procedure specific to Gaussian Processes. Bi et al.~\cite{bi2014learning}argue that the factors that determine the annotators' labeling performance, apart from the expertise, can also be the worker's dedication to the task, his/her default labeling judgement and sample difficulty. However, they do not go further in determining the optimal subset of annotators. On the other hand, Li and Liu~\cite{li_liu_2015} propose a combinatiorial procedure to find this optimal subset. However, such combinatorial procedures are very resource-intensive, especially for a large initial set of annotators.

@@

We design a methodology that both selects an optimal set of annotators from a set of training data and automatically determines the weights of the selected annotators. By forcing sparsity on our annotator weight vector, we get an accurate classification system that works well even with annotations from a small set of highly competent annotators.

In summary, we make the following contributions:
\begin{itemize}
\item We propose a methodology for selecting an optimal subset of workers in a classification system with multiple labelers
\item We create a procedure for selecting and determine weights of labelers independent of type machine learning methods selected for modeling the annotating process
\item Using a set of experiments with real and synthetic data, we show that we can obtain accurate classification system with the learned annotator subset.
\end{itemize}

@@

Our methodology is based on the scenario drawn on the \autoref{fig:scheme}. This scheme shows a common situation when a server gets labells from multiple annotators.

@@

In training time, server forwards data to all the clients in order to retrieve annotations. Using the data annotated from a large set of clients, server trains a machine learning model that jointly learns the expertise of all the clients while minimizing the quadratic loss with respect to the ground truth.

@@

\begin{enumerate}
\item minimizing the error with respect to the ground truth
\item minimizing the error with respect ot the client annotations
\item making the client weight vector sparse 
\end{enumerate}

For example, the cost function for logistic regression looks like the following:

\begin{equation}
\begin{split}
 L(w,v)= -\frac{1}{N} \sum_{n=1}^N (\hat{y}_n \log \sum_{i=1}^M v_i y_{ni} + (1-\hat{y}_n) \log(1-\sum_{i=1}^M v_i y_{ni})) + \\ \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\lambda \mid v \mid
\end{split} 
\end{equation}

\begin{equation}
\sum_{i=1}^{M} v_i = 1 
\end{equation}

where $ \hat{y}_n $ are ground truth labels for all training samples, $ y_{ni} $ are labels retrieved by querying  current client models, $z_{ni}$ are training labels from annotator clients, $v_i$ is a normalized vector of weights for all the clients. The final loss function consists of a crossentropy calculation for the empirical loss with respect to the ground truth, the quadratic loss with respect to the client annotations weighted by a parameter $\phi$ and the 1-norm of the parameter vector $v$ weighted by another parameter $\lambda$.

@@

\begin{equation} \frac{\partial L}{\partial v_i} =  -\frac{1}{N} \sum_{n=1}^N (y_n \frac{1}{\sum_{i=1}^M v_i y_{ni}} + (1-\hat{y_n}) \frac{1} {1-\sum_{i=1}^M v_i y_{ni}} (-y_n) 
\end{equation}
The expression for the gradient $ \frac{y_n}{w_j}$ depends on the concrete method used to model the annotators.

For logistic regression we directly define the cost for the errors with respect to ground truth as crossentropy. However, with minimal changes our framework can also be usable with other machine learning methods. For example, we tested our methodology with neural networks as well. These methods, however, need to be convex or have a convex relaxation in order to still make a gradient-based optimization procedure work well.

In order to still use gradient descent as a convex optimization method with the $L_1$ norm, we use the method of subgradient when updating the value of $v$. The update that we derived is the following:
\begin{equation}
 v= v + S_{\lambda} (v-t_s \frac{\partial L}{\partial v}, \lambda)
\end{equation}
where
\begin{equation}
S_{\lambda} (k, \lambda) = 
\left\{
	\begin{array}{ll}
		k-\lambda  & \mbox{if } k > \lambda  \\
		k+\lambda & \mbox{if } k < -\lambda \\
		0 & \mbox{otherwise}
	\end{array}
\right.
\end{equation}

@@

Multiclass classification:

\begin{equation}
 L(w,v)= -\frac{1}{N} \sum_{n=1}^N \sum_{k=1}^K (\hat{y}_{nk} \log \sum_{i=1}^M v_i y_{nik}) + \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\lambda \mid v \mid
\end{equation}

Full expertise model:

\begin{equation}
\begin{split}
 L(w,v)= -\frac{1}{N} \sum_{n=1}^N (\hat{y}_n \log \sum_{i=1}^M v_{ni} y_{ni} + (1-\hat{y}_n) \log(1-\sum_{i=1}^M v_{ni} y_{ni})) + \\ \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\sum_{n=1}^{N}\lambda \mid v_n \mid
\end{split} 
\end{equation}

@@

\subsection{Datasets}
In order to evaluate our approach, we used one dataset with real-life data and one synthetic dataset. We obtained a real-life dataset from the authors of a related paper about learning to predict from crowdsourced data by Bi et al.~\cite{bi2014learning}. This dataset consists of results from an experiment with 21 workers at the \textit{Amazon Mechanical Turk}. These workers executed 10 tasks, all of the tasks being binary classification of images between two types of dogs. Since each worker has different expertise, the challenge is to determine the subset of highly competent workers and combine their abilities for a superior classification system.

Synthetic dataset was created using a method similar to the one used by Huang et al.~\cite{xiao2013learning}. The dataset consists of 1-dimensional data, random points between $0$ and $2\pi$, also labeled using randomization. We generate a ground truth based on random noise superpositioned over a nonlinear function:

\begin{equation}
f_n = 10\sin(6x)\sin(x/2)
\end{equation}

In order to get proper labels for our data instead of this nonlinear function, we need to additionaly put the values of the function $f_n$ through a sigmoid function:

\begin{equation}
\sigma(x) = \frac{1}{1+e^t}
\end{equation}

Finally, we take the values higher than $0.5$ as $1$, while the other values correspond to label $0$.

@@

In the beginning, we examine the changes of the loss function when increasing the parameter $\lambda$ from 0 to 1, thereby adding the weight to the sparsity part of the loss function. The results of this test are displayed on Figure~\ref{}.

Furthermore, we look at how the value of the loss function changes while the optimization is running. We plot the loss against the number of iteration, which shows the steep decline of the loss value. This shows that using our gradient-based procedure we can get fast to a very accurate classification system. The graph is shown on Figure~\ref{}.

\subsection{Synthetic data}

In this subsection we show the classification performance on our synthetically generated data.

@@

% ---- Bibliography ----
%
\bibliographystyle{plain}
\bibliography{draft} % use this
%\clearpage
%\addtocmark[2]{Author Index} % additional numbered TOC entry
%\renewcommand{\indexname}{Author Index}
%\printindex
%\clearpage
%\addtocmark[2]{Subject Index} % additional numbered TOC entry
%\markboth{Subject Index}{Subject Index}
%\renewcommand{\indexname}{Subject Index}
%\input{subjidx.ind}
\end{document}

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern

\usepackage{subcaption}
\usepackage{color}
%\usepackage{subfig}%[caption=false,font=footnotesize]

@@

\section{Introduction}
Supervised machine learning depends on our ability to gather a large set of labeled data. In a high number of scenarios, we can benefit from labeled data gathered by querying multiple separate sources with varying reliability and join this data into one machine learning model. However, many times we are limited in the number of data sources that we are allowed to use, because of budget constraints, limited processing power or insufficient communication bandwidth. This causes the need for a framework that would enable resource-efficient combination of labeled data from multiple sources.

First important fitting scenario to this framework is crowdsourcing. When using crowdsourcing systems such as Amazon Mechanical Turk, we can gather labelled data from multiple annotators. After gathering this training data, we need to join the labelling decisions and train one machine learning system while leveraging all of the annotations. Relying on a concept of \textit{wisdom of crowds}, we can assume that if we gather data from a sufficient number of people, on average most of them will give correct labelling decision. However, gathering data from a large amount of annotators can be expensive, as workers on Amazon Mechanical Turk require payment for their work. Therefore we would benefit from a procedure of prior determination of the annotators' expertise. In this case, the joining process can be optimized by estimating the expertise of the annotators, selecting the optimal subset and determining weights of the annotators' labels. Moreover, in a rising area of \textit{Internet of Things} (IoT) we can have similar problems. In IoT systems, a large number of sensors provide information about the environment. Communication among multiple sensors can be costly in terms of processing power. Furthermore, this communication can be done among sensors that are limited with power consumption constraints. Furthermore, in intrusion detection it is useful to use results from multiple anomaly detection systems. For example, although there are multiple antivirus programs that can label software as benign or malicious programs of certain type, only a minority of the programs contain updated labels for newest malware families and classify malware with high granularity. Therefore it would be useful to find a subset of label sources that are reliable enough to be used for malware triage purposes.

There are multiple papers that deal with estimating the expertise of labelers when using \textit{wisdom of crowds}. For example, Welinder et al. \cite{welinder2010multidimensional} determine groups of annotators with similar expertise, and find particular phenomena such as "schools of thought" and specialist experts in certain types of image labeling tasks. Tian and Zhu~\cite{tian2012learning} investigate the annotator behavior in executing more complex labeling tasks, such as answering if an image is beautiful or if it contains a car. Zhang et al.~\cite{zhang2013learning} propose a method to filter out novice labelers. However, their method is specific to Gaussian Mixture Models. Rodrigues et al.~\cite{rodrigues2014gaussian} determine a similar procedure specific to Gaussian Processes. Bi et al.~\cite{bi2014learning} argue that the factors that determine the annotators' labeling performance, apart from the expertise, can also be the worker's dedication to the task, his/her default labeling judgement and sample difficulty. However, they do not go further in determining the optimal subset of annotators. On the other hand, Li and Liu~\cite{li_liu_2015} propose a combinatiorial procedure to find this optimal subset. However, such combinatorial procedures are very resource-intensive, especially for a large initial set of annotators.

@@

We design a methodology that both selects an optimal subset of annotators from a set of training data and automatically determines the weights of the selected annotators. By forcing sparsity on our annotator weight vector during optimization, we get an accurate classification system that works well even with annotations from a small set of highly competent annotators. This gives us a low-budget supervised learning strategy, which is not specific to any particular learning algorithm.

In summary, we make the following contributions:
\begin{itemize}
\item We propose a methodology for selecting an optimal subset of workers in a classification system with multiple labelers
\item We create a procedure for selecting and determining weights of labelers independent of the type of machine learning methods selected for modeling the annotating process
\item Using a set of experiments with both real and synthetic data, we show that we can obtain an accurate classification system with the optimized annotator subset.
\end{itemize}

@@

Our methodology is based on the scenario drawn on the \autoref{fig:scheme}. This scheme shows a common situation when a server acquires labels from multiple annotators.

@@

In training time, server forwards the training inputs to all the annotators in order to retrieve labels. Using the data labeled from a large set of annotators, the server trains a machine learning model that jointly learns the expertise of all the clients while minimizing the quadratic loss with respect to the ground truth.

@@

\begin{enumerate}
\item minimizing the cost with respect to the ground truth labels on the training set
\item minimizing the cost with respect ot the client annotations
\item making the client weight vector sparse (select a small subset of clients)
\end{enumerate}

The cost function for cross-entropy loss on ground truth, used in all our experiments, looks like the following:

\begin{equation}
\begin{split}
 L(w,v)= -\frac{1}{N} \sum_{n=1}^N (\hat{y}_n \log \sum_{i=1}^M v_i y_{ni} + (1-\hat{y}_n) \log(1-\sum_{i=1}^M v_i y_{ni})) + \\ + \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\lambda \mid v \mid
\end{split} 
\end{equation}
subject to:
\begin{equation}
 \sum_{i=1}^{M} v_i = 1 
\end{equation}

\noindent where $ \hat{y}_n $ are ground truth labels for all training samples, $ y_{ni} $ are labels retrieved by querying  current client models, $z_{ni}$ are training labels from annotator clients, $v_i$ is a normalized vector of weights for all the clients. The final loss function consists of a crossentropy calculation for the empirical loss with respect to the ground truth, the quadratic loss with respect to the client annotations weighted by a parameter $\phi$ and the $L_1$-norm of the parameter vector $v$ weighted by another parameter $\lambda$.

@@

\begin{equation} \frac{\partial L}{\partial v_i} =  -\frac{1}{N} \sum_{n=1}^N (y_n \frac{1}{\sum_{i=1}^M v_i y_{ni}} + (1-\hat{y_n}) \frac{1} {1-\sum_{i=1}^M v_i y_{ni}} (-y_n) 
\end{equation}
The expression for the gradient $ \frac{\partial y_n}{\partial w_j}$ depends on the concrete method used to model the annotators.

For logistic regression we directly define the cost for the errors with respect to ground truth as crossentropy. However, with minimal changes our framework can also be usable with other machine learning methods. For example, we tested our methodology with neural networks as well, using a perceptron with one hidden layer. These methods, however, need to be convex or have a convex relaxation in order to still make a gradient-based optimization procedure work well.

In order to still use gradient descent as a convex optimization method with the $L_1$ norm, we use iterative soft-thresholding~\cite{bredies2007iterative}, a type of generalized gradient descent optimization when updating the value of $v$. The update that we derived is the following:

\begin{equation}
 v= v + S_{\lambda} (v-t_s \frac{\partial L}{\partial v}, \lambda)
\end{equation}
where
\begin{equation}
S_{\lambda} (k, \lambda) = 
\left\{
	\begin{array}{ll}
		k-\lambda  & \mbox{if } k > \lambda  \\
		k+\lambda & \mbox{if } k < -\lambda \\
		0 & \mbox{otherwise}
	\end{array}
\right.
\end{equation}

@@

In case of multiclass classification, we define the loss function in a slightly different way, to accomodate with the fact that we need a label vector instead of scalar value:

\begin{equation}
 L(w,v)= -\frac{1}{N} \sum_{n=1}^N \sum_{k=1}^K (\hat{y}_{nk} \log \sum_{i=1}^M v_i y_{nik}) + \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\lambda \mid v \mid
\end{equation},

subject to
\begin{equation}
 \sum_{i=1}^{M} v_i = 1 
\end{equation}

This time, $\hat{y}$ is a matrix and $y$ is a three-dimensional tensor. The optimization is done in a similar way, using projected gradient descent. We minimize the cross-entropy w.r.t the ground truth, as well as the squared difference of ground truth and annotator decision, while keeping the annotator vector sparse.

%Full expertise model:
%
%
%\begin{equation}
%\begin{split}
% L(w,v)= -\frac{1}{N} \sum_{n=1}^N (\hat{y}_n \log \sum_{i=1}^M v_{ni} y_{ni} + (1-\hat{y}_n) \log(1-\sum_{i=1}^M v_{ni} y_{ni})) + \\ \psi \sum_{i=1}^{M} (y_{ni} - z_{ni})^2+\sum_{n=1}^{N}\lambda \mid v_n \mid
%\end{split} 
%\end{equation}

@@

\subsection{Datasets}
In order to evaluate our approach, we use one dataset with real-life labelers and one with synthetic labelers.

\subsubsection{Real-Life Labelers}
We obtained a real-life labeler set from the authors of a related paper about learning to predict from crowdsourced data by Bi et al.~\cite{bi2014learning}. This dataset consists of results from an experiment with 21 workers at the \textit{Amazon Mechanical Turk}. These workers executed 10 tasks, all of the tasks being binary classification of images. Since each worker has different expertise, the challenge is to determine the subset of highly competent workers and combine their abilities for a superior classification system.

\subsubsection{Synthetic Labelers}
In order to further examine the performance of our method, we also use a semi-synthetic data based on the well-known MNIST dataset~\cite{lecun1998mnist} for digit recognition with around 70000 images of digits. We use images from this data without change and we take labels for images as ground truth. However, we synthetically create annotators by training multiple classifiers with random 0.5\% of images from the original set. This gives us annotators that have expertise in labeling only a subset of images, that on average have 81\% accuracy on a holdout test set. We use this set of annotators to test if we can surpass the performance of a single particular annotator by combining a minority of them.

Furthermore, we look at how the value of the loss function changes while the optimization is running. We plot the loss against the number of iteration, which shows the steep decline of the loss value. This shows that using our gradient-based procedure we can get fast to a very accurate classification system. The graph is shown on \autoref{fig:optimization}.

\begin{figure}[!htb]	
    \centering

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/loss_iteration_logreg}
        \caption{Loss minimization}

    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/v_iteration_logreg}
        \caption{Reduction of annotator subset}

    \end{subfigure}
    \caption{Training process}
    \label{fig:optimization}
\end{figure}

\subsection{Performance per task}

In the following figures we examine the breakdown of classification error rate per labeling task. We can notice on \autoref{fig:pertask} that the performance of our classifier ensemble has a high variation, both when using a neural network and logistic regression.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/task_logreg}
        \caption{A gull}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/task_mlp}
        \caption{A gull}
    \end{subfigure}
    \caption{Performance per task}
    \label{fig:pertask}

\end{figure}

\subsection{Comparison with baseline ensemble methods}

Furthermore, we compare the performance of our scheme with annotator selection to some baseline ensemble methods, such as averaging annotator values and taking a majority vote. %Moreover, we also compare our performance with a system that randomly chooses annotators. 
This comparison is shown on Figure~\ref{fig:baseline}.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/comp_average}
        \caption{A gull}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/comp_average_mlp}
        \caption{A gull}
    \end{subfigure}
    \caption{Comparison with baseline}
    \label{fig:baseline}
\end{figure}

\subsection{Comparison of sparse and dense weight vectors}

In this subsection we test the performance penalty of enforcing sparsity on the annotator weight vectors. We evaluate our assumption that a low number of high-quality annotators can together incur high performance. This turns out to be true, according to our results, shown on \autoref{fig:sparsity}.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{A gull}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{A gull}
    \end{subfigure}
    \caption{Effect of sparsity on model accuracy}
    \label{fig:sparsity}
\end{figure}

@@

In the beginning, we examine the changes of the loss function when increasing the parameter $\lambda$ from 0 to 1, thereby adding the weight to the sparsity part of the loss function. The results of this test are displayed on \autoref{fig:sensitivity}.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/accuracy_lambda}
        \caption{A gull}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/accuracy_lambda_mlp}
        \caption{A gull}
    \end{subfigure} 
  \caption{Sensitivity to the hyperparameter}
  \label{fig:sensitivity}
\end{figure}

\subsection{Accuracy on synthetic data}

In this subsection we show the classification performance on our synthetically generated data. We show on \autoref{fig:synthetic}.that a small number of efficient labeling experts can give a much higher performance than individual labelers.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{Logistic Regression}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{Neural Network}
    \end{subfigure}
  \caption{Accuracy for MNIST data}
  \label{fig:synthetic}
\end{figure}

\subsection{Comparison of sparse and dense vectors}

In the next figures we show the breakdown of performance for sparse selection w.r.t the selection with no sparsity enforcement. On \autoref{fig:sparsity_mnist} it is noticeable that the error rate is uniform.

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{Logistic Regression}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{Neural Network}
    \end{subfigure}
  \caption{Performance per digit}
  \label{fig:sparsity_mnist}
\end{figure}

\subsection{Size of the annotator set}

In the next group of plots we show how the size of the annotator subset grows with the initial number of annotators. This is shown on  \autoref{fig:subset}

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{A gull}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/plot_mlp}
        \caption{A gull}
    \end{subfigure}
  \caption{Change of size of the annotator groups}
  \label{fig:subset}
\end{figure}

@@

In this section, we present the methodology we utilize for the continuous authentication. To achieve this we use four stages: $(i)$~experiment design, $(ii)$~data collection, $(iii)$~feature engineering and selection, and $(iv)$~online learning. In the following we present the threat model and each stage of building our authentication scheme.

\subsection{Threat Model}\label{ssec:threat}
In our methodology we consider the situation where we train our model incrementally using a valid user and then, in the test time, have both a valid user and an attacker trying to use the device. Both the attacker and the valid user leave a trace of their activity in the form of touchscreen interaction, accelerometer, and rotation sensor values. The goal of our system is to detect the attacker's behavior as a significant deviation from the model based on the valid user, while letting the valid user use the device without unnecessary interruption.  Therefore, we consider our problem as an example of unsupervised anomaly detection problems.

\subsection{Experiment Design}

To gather user-specific data, we design a two-part experiment focusing on different typical activities on the smartphone. For the first part, we focus on the simple activities consisting of a small set of touchscreen gestures. Therefore, we create a mobile app for Android 5.0, specifically designed and implemented for this purpose. We instruct users to execute actions on a smartphone based on the following gestures: swipe, keystroke, double tap, and pinch. For the second part, we leverage mobile apps that are typically used by smartphone users. We instruct users to perform some actions on popular smartphone apps in order to collect their sensor value patterns when using these applications.

\medskip \noindent \textbf{\emph{Requested Tasks in our Mobile App.}}
We request users to perform a series of tasks so we can train our model. A potential future app, delivered directly from the smartphone vendor, could offer the option for users to train their smartphones with their behavior, if they wish to enable the continuous authentication feature. Another possibility is for the user to perform the tasks presented in our app under the hood. This means that the phone does not need to execute an application, where it requests input from the user, but it can stealthily collect data and train the models. The tasks that we use for collecting continuous authentication data in our app are the following:

\begin{enumerate}
%
\item \emph{Image Finding.} We show the users 20 pictures in a row. By using horizontal and vertical scrolling they should find one particular image given in the instructions. The correct picture is then selected by a double tap.
%
\item \emph{Text Reading}. The user is asked to first read a text. To read the whole text they have to scroll down (or perhaps scroll up in case they miss some information). After reading the text, they have to answer five questions, which are displayed below the text.
%
\item \emph{Zooming.} We display a picture together with a corresponding question about a specific detail. The picture is either zoomed in or zoomed out so that the users could not find the detail without zooming the picture to the appropriate screen size. %We repeat this task ten times with different pictures and details.
%
\item \emph{Message Composition.} In this task, we ask users to enter a message and a recipient. %The data is given on the task description handed out to participants.
This way, we emulate users' behavior when composing an email or a SMS message.
%
\item \emph{PIN.} We ask users to enter a four-digit PIN five times in a row. The repetition makes the user adding the PIN more natural, similr to what happens in real-life.
%
\item \emph{Phone Call.} After having entered a given phone number correctly, users have to press the call button. In total three phone numbers are given, making the users repeat the task for each number.
%
\end{enumerate}

We design these tasks taking into account the most often performed activities on a smartphone. Each task is repeated at least ten times per participant which lowers the risk of sudden change of behavior because of external conditions (e.g. stress).

\medskip \noindent \textbf{\emph{Tasks with Built-in Apps.}}
We execute the next series of tasks using the apps provided by the mobile devices themselves. This makes the behavioral traces of users more natural, as they regularly use this kind of apps and can perform the tasks more smoothly.

\begin{enumerate}
%
\item \emph{Browsing.} We ask users to search on Google and read three specific Wikipedia articles. To verify that users read the correct articles, they had to type a correct answer to three questions about the corresponding paragraphs.
%
\item \emph{Taking a Selfie.} Users are asked to take a selfie with the smartphone camera.
%
\end{enumerate}

\subsection{Data Collection}

We used the \textit{Samsung Galaxy S4} smartphones with Android 5.0 to execute the above described experiments with 28 participants in order to capture the discriminative features in user activity.

\medskip \noindent \textbf{\emph{Motion Events.}}
The Android API delivers a sequence of motion events to the currently active application each time a user touches the screen. The touch sequence is started by a down event followed by a variable number of move events and an up event closing the sequence. Each motion event can be identified by its timestamp and its $x$ and $y$ coordinate. Timestamps are reported in \textit{ms} since the last reboot of the device. Besides, major and minor axis of the ellipse making up the touch contact are reported as well as the approximate size of the contact area. The size is calculated as mean of major and minor of the ellipse axes. To distinguish different types of events, we label them with an event ID. To support multitouch, each event is additionally labeled with a pointer ID to distinguish events caused by different fingers on the screen.

\medskip \noindent \textbf{\emph{Custom Keyboard.}}
For security reasons our version of Android does not allow to capture keyboard input of the standard keyboard. Therefore a custom keyboard is developed that resembled the Samsung keyboard, which was running as an Android service. Users are instructed to choose the custom keyboard when opening the app such that no other keyboard could accidentally be used. Apart from the motion events described in the previous subsection, the keycode of each key hit is captured.

\medskip \noindent \textbf{\emph{Sensor Service.}}
In addition to touch data, sensor data from built-in accelerometer and gyroscope (rotation sensor) is harvested. Data is polled at a rate of five times a second. The accelerometer measures acceleration in \textit{m/s2}, while the gyroscope measures rate of rotation in \textit{rad/s} along $x$, $y$ and $z$ axis of the phone. Gathering sensor data is done by running an Android service, same as in case of the custom keyboard.

\medskip \noindent \textbf{\emph{Input Device Driver.}}
Android only allows capturing touch events from within a custom application. Therefore, it is not possible to gather motion events from the Android API while using standard applications for the second part of the experiment. Nevertheless, it is possible to directly read from the input device driver under /dev/input/eventX, where X is a number identifying the input device. Whenever the user leaves the app, touch events are automatically read from the input driver: it reports a stream of single values not processed to higher level motion events as in case of the Android API. Each value is reported with its timestamp, a label identifying the type of input, and its actual hexadecimal value. The types of inputs are: $(i)$~$x$ and $y$ coordinates, $(ii)$~major and minor ellipse axis, and $(iii)$~a pointer counter that is increased each time a finger hits the screen. Since the devices used in this study are not capable of directly recording the user's pressure when touching the screen, we could not evaluate this feature. Although we could not measure this pressure directly, acceleration in $z$-direction together with touch area size gives information about the device's response to the force applied to the screen.

\subsection{Feature Engineering and Selection}

Using ideas from related work, we capture a large range of spatio-temporal features to determine which subset is the most useful. There are 59 features from swipe gestures and 51 features from keystrokes. The swipe feature set contains data that characterizes the trajectory of the swipe motion, the velocity of movement, and size of the trace. On the other hand, the keystroke features contain the similar velocity and stroke size patterns as well as keycode and offsets to the key center. Our feature set is based on the works of Velten et al~\cite{velten2015user}, Frank et al.~\cite{Frank13}, and Buschek et al.~\cite{buschek2015}.

Before training the model we execute a simple feature selection process. Our feature selection is based on the sequential feature selection procedure. According to the sequental feature selection algorithm~\cite{ruckstiess2011sequential}, in each round $k$ the feature is added to the feature subset $Y_k$ that yields highest classification accuracy $J( Y_k )$. Starting with the empty set, we compare classification accuracy for each feature that has not already been added to our feature subset in previous rounds. This has been done for 30 differently sampled user combinations. The feature that is chosen by most of the validation tests is added. If features yield same results one is randomly selected. We execute this procedure for the keyboard and swipes data separately.

\subsection{Online Learning}

This section introduces the notation used throughout the paper. A summary of notation can be found in Table~\ref{tbl:notation}.

\begin{table}
\centering
\begin{tabular}{lll}
	\toprule
	\textbf{Notation}       &   & \textbf{Explanation}                               \\ \midrule
	$x^{t}$                 & : & Incoming sample at time t                          \\
	$x_{i}$                 & : & $i$-th support vector                               \\
	$K(\cdot,\cdot)$        & : & RBF-kernel                                         \\
	$\gamma$                & : & Kernel parameter                                   \\
	$f_{t}$                 & : & Model at time $t$                                  \\
	$\alpha_{i}$            & : & Weight for $i$th support vector                    \\
	$y^{t}$                 & : & Correct class label at time $t$                    \\
	$\hat{{y}}^{t}$         & : & Predicted class label at time $t$                  \\
	$\ell(f;(x^{t},y^{t}))$ & : & Loss function                                      \\
	$B$                     & : & Budget size limiting the number of support vectors \\
	$X$                     & : & Set of support vectors                             \\
	$margin$                & : & Distance from decision boundary                    \\ \bottomrule \\
\end{tabular}
\caption{Summary of notation.}
\label{tbl:notation}
\end{table}

\begin{comment}
%
\begin{minipage}{0.3\textwidth}
	\vspace{\baselineskip}
	$x^{t}$\\
	$x_{i}$\\
			$K(\cdot,\cdot)$\\
			$\gamma$\\
			$f_{t}$\\
			$\alpha_{i}$\\
			$y^{t}$\\
			$\hat{{y}}^{t}$\\
			$\ell(f;(x^{t},y^{t}))$\\
			$B$\\
			$X$\\
			$margin$\\

\end{minipage}
\begin{minipage}{0.7\textwidth}
	\vspace{\baselineskip}
	Incoming sample at time t\\
	$i$th support vector \\
	RBF-kernel \\
	Kernel parameter\\
	Model at time $t$\\
	Weight for $i$th support vector\\
	Correct class label at time $t$\\
	Predicted class label at time $t$\\
	Loss function\\
	Budget size limiting the number of support vectors\\
	Set of support vectors\\
	Distance from decision boundary\\

\end{minipage}
\end{comment}

Xu et al.~\cite{xu2014towards} have shown that user behavior is not stable over time. To be able to continuously authenticate a user, we need our model to adapt to the current behavior. For this reason batch learning techniques are inappropriate in this setup, as they rely on training data being available in advance. They are not able to adapt to changes occurring after the initial training phase. Instead, the model should change whenever needed as well as be constantly evaluated in characterizing current user behavior. Online machine learning therefore constitutes a logical solution.

Algorithm~\ref{alg:online} introduces the basic online learning principle. Instead of training only once on a larger data set, online learners learn incrementally whenever new data is available. As they are only presented one sample at a time, online learners need to efficiently deal with limited information: they continuously adapt their learned decision boundary to the incoming stream of data. On receiving a new sample $x_{t}$ at time t, they predict its class label based on the current model $f_{t}$. Afterwards, they compare the retrieved label to the true class label given after the prediction and update the model. Many of the models are built in a nonparametric way to become more flexible.

 \begin{algorithm}[H]
 	\caption{Online Learning}
 	\label{alg:online}
 	\begin{algorithmic}[1]
 		\For{t = 1,2, ... }
 		\State receive instance $x^{t}$
 		\State predict label: $\hat{{y}}^{t} = sign(f_{t}(x^{t}))$
 		\State receive correct label $y_{t}$;
 		\State update model;
 		\EndFor
 	\end{algorithmic}
 \end{algorithm}

However, nonparametric online methods suffer from unlimited growth of the model over time. As we concentrate on bounding memory requirements in this study due to the resource-constrained environment on the device, we aim to to limit the model size. The next section introduces an online algorithm on a budget that is able to limit the number of support vectors and meet this requirement. For comparison, online learning algorithms with unlimited budget will also be tested 
in Section~\ref{sec:evaluation}.
